The function--sheaf dictionary for smooth commutative group schemes over finite fields

### 2014-04-03 dictionary.tex
Cut out that really long annoying proposition, which was based on a false premise (the existence of global sections) and replaced it with two news lemmas.
Still working on the section on etale group schemes.

In order to relate $\QCiso{G}$ to $G(k)^*$, we need a simple result relating duals, invariants and coinvariants.

@@

\begin{lemma}\label{prop:etale-trace}
If $G$ is an \'etale commutative group scheme over $\Fq$
then for every the $\chi \in G(\Fq)^*$ there is a visible quasicharacter sheaf $\qcs{L}$ on $G$ such that $\trFrob{\qcs{L}} = \chi$.
In particular, Trace of Frobenius
\[
\TrFrob{G} : \QCiso{G} \to G(\Fq)^*,
\]
is surjective.
\end{lemma}

@@

\begin{proposition}\label{prop:etale-fg}
  If $G$ is an \'etale commutative group scheme over $\Fq$ and
  $G(\bFq)$ is finitely generated,
  then
  \[
  \QCiso{G} \iso \Hh^2(G(\bFq)\rtimes \Weil{}, \EEx).
  \]
  Moreover, this isomorphism is functorial
  in the sense of Proposition~\ref{prop:functorialG}.
\end{proposition}

  We may associate a $1$-cocycle $\tau_\qcs{L}: \Weil{}\to G(\bFq)^*$ to each $\qcs{L} \in \QC(G)$
  so that
  \[
  \tau_\qcs{L}(\Frob{})(x) = \trFrob{\qcs{L}}(x)
  \]
  for all $x \in G(\Fq)$ and the map
  \begin{align*}
  \QCiso{G} &\to \Hh^1(\Weil{}, G(\bFq)^*),\\
  [\qcs{L}] &\mapsto [\tau_\qcs{L}],
  \end{align*}
  is an isomorphism.

\begin{proof}
  Suppose $\qcs{L}$ is a quasicharacter sheaf on $G$ and choose a global section $s$
  of $\qcs{L}$ using Lemma~\ref{lemma:section}.
   We will construct a group isomorphism between $\QCiso{G}$ and
  $\Hh^1(\Weil{}, G(\bFq)^*)$ using $s$, then prove independence from the choice of $s$.

  Starting from $\qcs{L}$, we define a cocyle $\tau_{\qcs{L}} \in Z^1(\Weil{},G(\bFq)^*)$ as follows.
  For $x \in G(\bFq)$, $\qcs{L}$ determines an isomorphism $\phi_x : \gqcs{L}_{\Frob{}(x)} \to \gqcs{L}_x$
  of one-dimensional $\EE$-vector spaces; define $\tau_\qcs{L}$ by
  \begin{equation}\label{t}
   \phi_{x}(s(\Frob{}(x))) = \tau_\qcs{L}(\Frob{})(x) s(x).
  \end{equation}
  Note that if $x$ is fixed by $\Frob{}$ then $\tau_\qcs{L}(\Frob{})(x)$ is just $\trFrob{\qcs{L}}(x)$, so the restriction of
  $\tau_\qcs{L}(\Frob{})$ to $G(k)$ is just $\trFrob{\qcs{L}}$.
  In general, \ref{qc.1} and \ref{qc.3}, together with \eqref{section}, guarantee that
  the scalar $\tau_\qcs{L}(\Frob{})(x)$ is non-zero.  Moreover,
  \ref{qc.2} forces
  $\tau_\qcs{L}(\Frob{})(x+y) = \tau_\qcs{L}(\Frob{})(x) \ \tau_\qcs{L}(\Frob{})(y)$
  for all $x,y \in G(\bFq)$.  We write $\cdot$ for pointwise multiplication of functions, and
  define $\tau_\qcs{L} : \Weil{} \to G(\bFq)^*$ recursively by
  \[
   \tau_\qcs{L}(\Frob{}^n) \ceq \tau_\qcs{L}(\Frob{})\cdot \,^{\Frob{}} \tau_\qcs{L}(\Frob{}^{n-1}),
  \]
  where $\Frob{}$ acts on $G(\bFq)^*$ through pre-composition.

  Although we used a global section $s$ in the definition of
  $\tau_\qcs{L}$, the class of $\tau_\qcs{L}$ does not depend on this
  choice. To see why, let $s_1$ and $s_2$ be two nontrivial global sections of
  $\qcs{L}$ and let $\tau_1$ and $\tau_2$ be the cocycles defined by $s_1$ and $s_2$.
  For each $x \in G(\bFq)$ define $a(x) \in \EEx$ by $s_2(x) = a(x) s_1(x)$.
  Then \eqref{section} and
  \ref{qc.2} together imply $a(x+y) = a(x)a(y)$, so
  $a\in G(\bFq)^*$.  The coboundary $\tau_0 : \Weil{} \to G(\bFq)^*$
  defined by $\tau_0(\Frob{}) \ceq \,^{\Frob{}} a \cdot a^{-1}$ satisfies $\tau_2 = \tau_0 \tau_1$,
  so $\tau_2$ and $\tau_1$ are cohomologous. Thus
  \begin{equation}\label{cohomologous}
  \begin{split}
    \obj \QC(G) &\to \Hh^1(\Weil{},G(\bFq)^*), \\
    \qcs{L} &\mapsto [\tau_\qcs{L}],
  \end{split}
  \end{equation}
  is independent of the choice of global sections made above.

  We now show that $[\tau_\qcs{L}]$ depends only on the isomorphism class of $\qcs{L}$.
  Let $\alpha : \qcs{L} \to \qcs{L'}$ be an isomorphism and
  let $\tau$ and $\tau'$ be any cocycles representing the classes in the
  image of $\qcs{L}$ and $\qcs{L'}$ under \eqref{cohomologous}; by construction,
  $\tau$ and $\tau'$ are defined by the choice of sections $s$ and $s'$.
  Now, $\varphi_x(s(x))$ is a
  non-zero scalar multiple of $s'(x)$; define $a: G(\bFq) \to \EEx$
  by $\alpha_x(s(x)) = a(x) s'(x)$.
  By the second part of \ref{qc.4},
  $a(x+y) = a(x)a(y)$, so $a \in G(\bFq)^*$. By the first part of
  \ref{qc.4},
  $\tau'(\Frob{}) = \tau(\Frob{}) \cdot (\,^{\Frob{}^{-1}}a \cdot a^{-1})$,
  so $\tau$ and $\tau'$ are cohomologous. This shows that
  \begin{equation}\label{pich1}
  \begin{split}
    \QCiso{G} &\to \Hh^1(\Weil{},G(\bFq)^*),\\
    [\qcs{L}] &\mapsto [\tau_\qcs{L}]
  \end{split}
  \end{equation}
  is well-defined.

    Suppose that $\qcs{L}$ and $\qcs{L}'$ are two quasicharacter sheaves on $G$.
  If $s$ is a global section for $\qcs{L}$ and $s'$ for $\qcs{L}'$ then $s \otimes s'$
  is a global section for $\qcs{L} \otimes\qcs{L}'$.  The three versions of \eqref{t}
  for $\qcs{L}, \qcs{L'}$ and $\qcs{L} \otimes\qcs{L}'$ imply that
  $\tau_{\qcs{L} \otimes\qcs{L}'}(\Frob{}) = \tau_\qcs{L}(\Frob{}) \tau_{\qcs{L}'}(\Frob{})$,
  so \eqref{pich1} is a group homomorphism.

 To see that \eqref{pich1} is injective,
 suppose $[\tau_\qcs{L}] = [\tau_\qcs{L'}]$.
 Let $s$ and $s'$ be sections appearing in the definition of $\tau_\qcs{L}$ and $\tau_\qcs{L'}$.
 Since $[\tau_\qcs{L}] = [\tau_\qcs{L'}]$,
 it follows that $\tau_\qcs{L}$ and $\tau_{\qcs{L'}}$ are cohomologous,
 so there is some $a\in G(\Fq)^*$ such that
 $\tau'(\Frob{}) = \tau(\Frob{}) \cdot (\,^{\Frob{}}a \cdot a^{-1})$.
 Define $\alpha : \qcs{L}\to \qcs{L'}$ by $\alpha_x(s(x)) \ceq a(x) s'(x)$ and $\alpha^{-1} : \qcs{L}'\to \qcs{L}$
 by $\alpha^{-1}_x(s'(x)) \ceq a(x)^{-1} s(x)$, for $x\in G(\bFq)$.
 It is straightforward to check that $\alpha$ satisfies \ref{qc.3}
 and that $\alpha^{-1}$ is its inverse. Thus $\qcs{L} \iso \qcs{L'}$ in $\QC(G)$ and \eqref{pich1} is injective.

 Finally, to show that \eqref{pich1} is surjective, pick $[\tau]\in \Hh^1(\Weil{}, G(\bFq)^*)$.
 For each $x\in G(\bFq)$, set $\gqcs{L}_x \ceq \EE$.
 Define $\mu_{x,y} : \gqcs{L}_{x+y} \to \gqcs{L}_x\otimes \gqcs{L}_y$ by
 $\mu_{x,y}(1) \ceq 1\otimes1$, for $x,y\in G(\bFq)$;
 clearly, $\mu$ satisfies \ref{qc.2}.
 Define $\phi_x : \gqcs{L}_{\Frob{}(x)} \to \gqcs{L}_x$ by $\phi_x(1) \ceq \tau(\Frob{})(x)$,
 for each $x\in G(\bFq)$.
 Then $\phi$ satisfies \ref{qc.3} because $\tau \in Z^1(\Weil{},G(\bFq)^*)$.
 Setting $\qcs{L} \ceq (\gqcs{L},\mu,\phi)$, we have that $[\tau]$ is the image of $\qcs{L}$
 under \eqref{pich1}, and thus \eqref{pich1} defines an isomorphism of groups.
\end{proof}

\begin{proposition} \label{prop:etale-iso}
 If $G$ is a commutative, \'etale group scheme over $\Fq$ and $G(\bFq)$ is finitely
 generated then
 \[
  \TrFrob{G} : \QCiso{G} \to G(\Fq)^*
 \]
 is an isomorphism of groups.
\end{proposition}
\begin{proof}
There is a natural isomorphism $\Hh^1(\Weil{}, G(\bFq)^*) \to (G(\bFq)^*)_{\Frob{}}$
defined by $[z] \mapsto z(\Frob{})$.  The proposition now follows directly
from Proposition~\ref{prop:etale} and Lemma~\ref{lem:dual-inv}.
\end{proof}

\section{Exactness of the component group sequence} \label{sec:restriction}

Consider the short exact sequence in the category of $\QC$-schemes
defining the component group scheme for $G$:
\begin{equation}\label{eq:pi0}
\begin{tikzcd}
0 \rar & G^0 \arrow{r}{\iota_0} & G \arrow{r}{\pi_0} & \pi_0(G) \rar & 0.
\end{tikzcd}
\end{equation}
Since $\pi_0(G)$ is also a $\QC$-scheme,
Proposition~\ref{prop:pullback} implies that \eqref{eq:pi0} defines the sequence of functors
\begin{equation}\label{eq:pi1}
\begin{tikzcd}
\QC(0) \rar & \QC(\pi_0(G)) \arrow{r}{\pi_0^*} & \QC(G) \arrow{r}{\iota_0^*} & \QC(G^0) \rar & \QC(0)
\end{tikzcd}
\end{equation}
and therefore, after passing to isomorphism classes, the sequence of abelian groups
\begin{equation}\label{eq:pi2}
\begin{tikzcd}
0 \rar & \QCiso{\pi_0(G)} \arrow{r}{\pi_0^*} & \QCiso{G} \arrow{r}{\iota_0^*} & \QCiso{G^0} \rar & 0.
\end{tikzcd}
\end{equation}
 Note that we found the groups $\QCiso{G^0}$
and $\QCiso{\pi_0(G)}$
in Sections~\ref{sec:connected} and \ref{sec:etale}.
In this section we show that \eqref{eq:pi2} is right exact.

@@

\begin{proposition}\label{prop:middleexact}
 The sequence
 \[
  \begin{tikzcd}
  0 \rar & \QCiso{\pi_0(G)} \arrow{r}{\pi_0^*} & \QCiso{G} \arrow{r}{\iota_0^*} & \QCiso{G^0} \rar & 0.
  \end{tikzcd}
 \]
 is exact at $\QCiso{G}$.
\end{proposition}

\begin{proof}
First note that $\iota_0^* \circ \pi_0^*$ is trivial by Proposition~\ref{prop:pullback}.
So it suffices to show that if $\qcs{L} = (\gqcs{L},\mu,\phi)$ is a quasicharacter sheaf on $G$
with $\qcs{L}\vert_{G^0} = (\EE)_{G^0}$ then $\qcs{L}$ is in the essential image of $\pi_0^*$.

>>>>

%In order to relate $\QCiso{G}$ to $G(k)^*$, we need a simple result relating duals, invariants and coinvariants.

@@

\begin{lemma}\label{prop:etale-trace}
If $G$ is an \'etale commutative group scheme over $\Fq$
then for every the $\chi \in G(\Fq)^*$ there is a visible quasicharacter sheaf $\qcs{L}$ on $G$ such that $\trFrob{\qcs{L}} = \chi$.
In particular, Trace of Frobenius
\[
\TrFrob{G} : \QCiso{G} \to G(\Fq)^*,
\]
is surjective, even when restricted to visible quasicharacter sheaves.
\end{lemma}

@@

\begin{proposition} \label{prop:etale-iso}
	If $G$ is an \'etale commutative group scheme over $\Fq$ 
	and if $G(\bFq)$ is finitely generated,
  	then
  	\[
  	\QCiso{G} \iso \Hh^2(G(\bFq)\rtimes \Weil{}, \EEx).
  	\]
  	Moreover, this isomorphism is functorial
  	in the sense of Proposition~\ref{prop:functorialG}.
\end{proposition}

\begin{proof}

\end{proof}

\section{Right-Exactness of the component group sequence} \label{sec:restriction}

Consider the short exact sequence in the category of $\QC$-schemes
defining the component group scheme for $G$:
\begin{equation}\label{eq:pi0}
\begin{tikzcd}
0 \rar & G^0 \arrow{r}{\iota_0} & G \arrow{r}{\pi_0} & \pi_0(G) \rar & 0.
\end{tikzcd}
\end{equation}
Since $\pi_0(G)$ is also a $\QC$-scheme,
Proposition~\ref{prop:pullback} implies that \eqref{eq:pi0} defines the sequence of functors
\begin{equation}\label{eq:pi1}
\begin{tikzcd}
\QC(0) \rar & \QC(\pi_0(G)) \arrow{r}{\pi_0^*} & \QC(G) \arrow{r}{\iota_0^*} & \QC(G^0) \rar & \QC(0)
\end{tikzcd}
\end{equation}
and therefore, after passing to isomorphism classes, the sequence of abelian groups
\begin{equation}\label{eq:pi2}
\begin{tikzcd}
%0 \rar & 
\QCiso{\pi_0(G)} \arrow{r}{\pi_0^*} & \QCiso{G} \arrow{r}{\iota_0^*} & \QCiso{G^0} \rar & 0.
\end{tikzcd}
\end{equation}
 Note that we found the groups $\QCiso{G^0}$
and $\QCiso{\pi_0(G)}$
in Sections~\ref{sec:connected} and \ref{sec:etale}.
In this section we show that \eqref{eq:pi2} is exact.

@@

\begin{proposition}\label{prop:middleexact}
 The sequence
 \[
  \begin{tikzcd}
%  0 \rar & 
 \QCiso{\pi_0(G)} \arrow{r}{\pi_0^*} & \QCiso{G} \arrow{r}{\iota_0^*} & \QCiso{G^0} \rar & 0.
  \end{tikzcd}
 \]
 is exact.
\end{proposition}

\begin{proof}
We already know the sequence is exact at $\QCiso{G^0}$. 
Here we show exactness at $\QCiso{G}$.
First note that $\iota_0^* \circ \pi_0^*$ is trivial by Proposition~\ref{prop:pullback}.
So it suffices to show that if $\qcs{L} = (\gqcs{L},\mu,\phi)$ is a quasicharacter sheaf on $G$
with $\qcs{L}\vert_{G^0} = (\EE)_{G^0}$ then $\qcs{L}$ is in the essential image of $\pi_0^*$.
